Visualizing the Independence Problem Using Extended Association and Mosaic Plots

### 2003-03-11 trunk/vcd/paper/draft.Rnw
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@222 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\SweaveOpts{engine=R,eps=FALSE}

@@

But this is not the only plausible way of aggregation of the Pearson
residuals. There are many thinkable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda(r_{ij})$, the sum of squares
is just one of them. Another functional which is very helpful for identifying
the cells which cause the potential dependence is the maximum of the absolute
values
\begin{equation} \label{eq:MaxAbs}
M = \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the culprits causing
the dependence can be easily identified. The critical values $c_\alpha$ can be
derived from the conditional permutation distribution of the table $(n_{ij})$.
Of course, the $\chi^2$ test \ref{eq:ChiSquare} can also be carried out as
a conditional test instead of using the unconditional asymptotic distribution.

@@

The two best known visualization techniques for independence in 2-way tables are
association plots and mosaic plots. Both are suitable to bring out departure of an
observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
The former focuses on the visualization of the Pearson residuals $r_{ij}$ (under
independence) while the latter primarily displays the observed frequencies
$n_{ij}$.

@@

\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.assocplot(goals, rot = 0, gp = gp.binary)
@
\caption{Association plot for the Bundesliga data.}
\label{fig:goals-assoc-classic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=6>>=
mosaicplot(t(goals), shade = TRUE, main = "")
@
\caption{Friendly mosaic plot for the Bundesliga data.}
\label{fig:goals-mosaic-classic}
\end{center}
\end{figure}

@@

As pointed out in Section \ref{sec:tests} the critical values for
the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
from the conditional permutation distribution. Instead of the hard-coded
values 2 and 4 the particular critical values, e.g., at the level
$\alpha = 0.9$ and $0.95$, for the table to be visualized could be
used, thus highlighting exactly those residuals who cause a (potential)
dependency within the table. At the moment these critical values
are derived in \texttt{vcd} by simulation of the underlying distribution.

As an alternative to the HSV colors the device-independent
HCL (Hue-Chroma-Luminance) color space \citep{vcd:Ihaka:2003} could be
used. The resulting association plot (using a critical value for
$\alpha = 0.95$) for the Bundesliga data can be seen in
Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
there is no evidence for rejecting the hypothesis of independence
in this table. Thus, there seems to be neither an advantage for
the home team nor for the away team.

\begin{figure}[htbp]
\begin{center}
<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.assocplot(goals, rot = 0)
@
\caption{Extended association plots for the Bundesliga data.}
\label{fig:goals-assoc-signif}
\end{center}
\end{figure}

To illustrate what a `postive' example looks like we give the association plot
for the famous Hair-Eye-Color data set cross-tabulating the hair and eye
color of 592 statistics students. It is easy to see the hypothesis
of independence is rejected at 95\% level by the maximum test (\ref{eq:MaxAbs})
due to higher frequencies of student with hair and eye color
black/brown, blond/blue and red/green and lower frequencies of
blond/brown and black/blue than would be plausible under independence.

@@

We have seen that this approach performs very well when the test statistic
which should be visualized is the maximum statistic (\ref{eq:MaxAbs}) but
it is difficult to extend this approach to general tests for independence,
in particular in the HCL space, where the range of chroma and luminance are
not independent. HSV space on the other hand has the advantage, that all
three dimensions hue, saturation and value have the range $[0,1]$ and can
be varied independently. Therefore, a different approach than the one
described above would be to use the hue for the sign of the residuals as
before, the saturation for the absolute size of the residuals and the
value as an indicator for the significance of some test statistic, only
using the full value when the overall test for independence rejects
the null hypothesis. The resulting mosaic plots under this paradigm
are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
be clearly seen that despite some large residuals there is no
evidence against independence for the Bundesliga, but that there the
null hypothesis has to be rejected for the Hair-Eye-Color data.

@@

The current implementation of association and mosaic plots in \textsf{R} suffers
from two main disadvantages: Firstly, it is not easy to recycle the
plots in conditioning plots or pairs plots (like mosaic matrices) as
they have been implemented using \textsf{R}'s base graphics engine.
The new implementation was written from scratch in \texttt{grid} offering
much more versatility amongst some other minor advantages and convenient
improvements. Secondly, the graphics parameters of the rectangles in
association and mosaic plots (like in almost all standard \textsf{R}
plots), like color and line type, cannot be specified for each cell
by the user. To overcome this, the user can specify in the current
implementation in \texttt{vcd} either arrays of graphics parameters
of the same dimensionality as the object that is plotted or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. Functions are provided for the shading schemes
described in the previous section.

@@

Two natural ways to use the visualization techniques described in the previous
sections would be to use (Trellis-like) conditioned plots or pairs plots
(like mosaic matrices) to visualize these more complex patterns of
independence.

>>>>

\SweaveOpts{engine=R,eps=FALSE}
\setkeys{Gin}{width=0.8\textwidth}

@@

But this is not the only plausible way of aggregation of the Pearson
residuals. There are many thinkable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda(r_{ij})$, the sum of squares
is just one of them. Another functional which is very helpful for identifying
the cells which cause the potential dependence is the maximum of the absolute
values
\begin{equation} \label{eq:MaxAbs}
M = \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the culprits causing
the dependence can easily be identified. The critical values $c_\alpha$ can be
derived from the conditional permutation distribution of the table $(n_{ij})$
which is the distributation generated by all possible tables with the same
row and column sums. Of course, the $\chi^2$ test (\ref{eq:ChiSquare}) can also be
carried out as a conditional test instead of using the unconditional asymptotic
distribution.

@@

The two best-known visualization techniques for independence in 2-way tables are
association plots and mosaic plots. Both are suitable to bring out departure of an
observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
The former focuses on the visualization of the Pearson residuals $r_{ij}$ (under
independence) while the latter primarily displays the observed frequencies
$n_{ij}$.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(goals, rot = 0, gp = gp.binary)
@
\caption{Association plot for the Bundesliga data.}
\label{fig:goals-assoc-classic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7>>=
mosaicplot(t(goals), shade = TRUE, main = "")
@
\caption{Friendly mosaic plot for the Bundesliga data.}
\label{fig:goals-mosaic-classic}
\end{center}
\end{figure}

@@

As pointed out in Section \ref{sec:tests} the critical values for
the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
from the conditional permutation distribution. Instead of the hard-coded
values 2 and 4 the particular critical values, e.g., at the level
$\alpha = 0.9$ and $0.99$, for the table to be visualized could be
used. Hence, exactly those residuals causing the (potential)
dependency within the table are highlighted. At the moment these critical values
are derived in \texttt{vcd} by simulation of the underlying distribution.

As an alternative to the HSV colors the device-independent
HCL (Hue-Chroma-Luminance) color space \citep{vcd:Ihaka:2003} could be
used. The resulting association plot (using a critical values for
$\alpha = 0.9$ and $0.99$) for the Bundesliga data can be seen in
Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
there is no evidence for rejecting the hypothesis of independence
in this table. Thus, there seems to be neither an advantage for
the home team nor for the away team.

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.assocplot(goals, rot = 0)
@
\caption{Extended association plots for the Bundesliga data.}
\label{fig:goals-assoc-signif}
\end{center}
\end{figure}

To illustrate what a `postive' example looks like we give the association plot
for the wll-known Hair-Eye-Color data set which cross-tabulates the hair and eye
color of 592 statistics students. It is easy to see that the hypothesis
of independence is rejected at 99\% level by the maximum test (\ref{eq:MaxAbs})
due to higher frequencies of students with hair and eye color
black/brown and blond/blue and lower frequencies of
blond/brown and black/blue than would be plausible under independence. Furthermore,
it can be seen that the residuals corresponding to the red/green and blond/hazel
cells are also significant at the 90\% level.

@@

We have seen that this approach performs very well when the test statistic
which should be visualized is the maximum statistic (\ref{eq:MaxAbs}) but
it is difficult to extend this approach to general tests for independence,
in particular in the HCL space, where the ranges of chroma and luminance are
not independent. The HSV space on the other hand has the advantage that all
three dimensions hue, saturation and value have the range $[0,1]$ and can
be varied independently. Therefore, a different approach than the one
described above would be to use the hue for the sign of the residuals as
before, the saturation for the absolute size of the residuals and the
value as an indicator for the significance of some test statistic, only
using the full value when the overall test for independence rejects
the null hypothesis. The resulting mosaic plots under this paradigm
are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
clearly be seen that despite some large residuals there is no
evidence against independence for the Bundesliga data, but that the
null hypothesis has to be rejected for the Hair-Eye-Color data.

@@

The current implementation of association and mosaic plots in \textsf{R} suffers
from two main disadvantages: Firstly, it is not easy to recycle the
plots in conditioning plots or pairs plots (like mosaic matrices) as
they have been implemented using \textsf{R}'s base graphics engine.
The new implementation was written from scratch in \texttt{grid} offering
much more versatility amongst some other minor advantages and convenient
improvements. Secondly, the graphics parameters of the rectangles in
association and mosaic plots (like in almost all standard \textsf{R}
plots), like color and line type, cannot be specified for each cell
by the user. To overcome this the current implementation in \texttt{vcd}
allows the user to specify either arrays of graphics parameters
of the same dimensionality as the object that is plotted or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. Functions are provided for the shading schemes
described in the previous section. Without this feature we would not be
able to specify the shading schemes described in the previous section
in such a flexible way.

@@

Two natural ways to use the visualization techniques described in the previous
sections would be to use (Trellis-like) conditioning plots or pairs plots
(like mosaic matrices) to visualize these more complex patterns of
independence.
